---
title: "Culpeper County Subdivision Analysis"
date: today
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    theme: flatly
    author: "Jonathan Knopf (jonathan@hdadvisors.net)"
    code-overflow: wrap
  # pdf:
  #   fig-pos: "H"
  #   mainfont: "Open Sans"
execute: 
  echo: false
  warning: false
  freeze: auto
#code-fold: true
#code-overflow: scroll
from: markdown+emoji
---

```{r}
#| label: setup
#| code-summary: "Setup code"

library(tidyverse)
library(sf)
library(janitor)
library(mapview)
library(hdatools)
library(scales)
#library(ggtext)
#library(ggridges)
library(kableExtra)
#library(reactable)
#library(patchwork)
library(formattable)

pal <- hda_pal_discrete()(6)

```

# Introduction

## Background

In early 2024, Culpeper County officials began considering potential changes to their subdivision ordinance, which governs how land can be divided into smaller parcels for development. According to County documents, these proposed changes aim to support Comprehensive Plan goals and address concerns about current regulations' effectiveness in preserving agricultural land and managing growth.

However, the proposed modifications---which change lot size minimums, clustering guidelines, and other variables---could potentially have the unintended consequence of restricting housing production and limiting overall supply in the county. Fully assessing the potential impacts of these reforms will help inform ongoing policy conversations.

## Objectives

The primary objectives of this project include:

* Determining the number of parcels potentially affected by proposed changes to the subdivision ordinance.
* Estimating the potential number of new subdivisions under current regulations versus proposed changes.
* Examining the scope of properties already reserved for conservation.

The secondary objective of this project is to assess how these proposed changes might affect the type and pace of residential growth in the county.

## Data

This analysis will use the following data.

### County parcel-level records

* Parcel ID / PID (join field for parcel shapefile)
* Acreage
* Zone
* Land Use Code / Description
* Neighborhood
* Parcels in Land Use Program by use category (Agricultural/Horticultural/Forest)
* Conservation easements

### County spatial data

* Parcel shapefile (with join field from tax parcel records) \*
* Subdivision boundary shapefile \*
* Future land use shapefile
* Urban Services Boundary shapefile
* Water sewer service area shapefile
* Zoning shapefile (if zoning not included in parcel records)

\* Necessary for analysis

### County reports 

Approved subdivision history:

* Annual data (1993-2023)
* Number of subdivisions and resultant lots for Major Divisions, Minor Divisions, 10 Acre Divisions, and Family Partitions
* Total number of lots and plats

A-1 and RA zoning breakdown:

* Total number of parcels and acreage zoned A-1 or RA
* Number of A-1 and RA parcels by size (21 categories from 0.1 to 500.0+ acres)
* Total number of parcels and acreage zoned R-1, R-2, R-3, R-4, or RR

### Other data

* U.S. Census Bureau Building Permit Survey
* FEMA National Flood Hazard Layer shapefile
* MLS sales records for new single-family homes 

## Policy

The primary policy sources used for this analysis are Culpeper County's Zoning Ordinance (last updated August 2023) and Subdivision Ordinance (last updated January 2024).

Descriptions of potential reforms are sourced from memos and presentations created by County staff for Board of Supervisors and Planning Commission meetings.

Additional reference materials include the current Culpeper County Comprehensive Plan (adopted 2023), as well as the following chapters in the Code of Ordinances:

* Ch. 6 - BUILDING REGULATIONS
* Ch. 8 - EROSION AND SEDIMENTATION CONTROL
* Ch. 12, Art. II - ASSESSMENT OF REAL ESTATE DEVOTED TO AGRICULTURAL, HORTICULTURAL AND FOREST USES (last updated 2017)
* APPENDIX D - CONSERVATION EASEMENT PROGRAM (last updated October 2023)

# Policy Analysis

::: {.callout-tip}
## Key Policy Takeaways

**1. Reduced development potential in rural areas**

The proposed increases in minimum lot sizes for A-1 and RA districts (from 5 and 3 acres to potentially 10-20 and 6 acres respectively) would significantly reduce the number of lots that could be created through subdivision in these areas. This could decrease the overall supply of new housing, potentially driving up prices.

**2. Shift towards clustered development**

Options 3, 4, and 5 introduce clustering provisions, which could maintain or even increase housing unit yield in some cases. However, this shift may lead to a different housing product (smaller lots in cluster arrangements) that could affect market demand and pricing.

**3. Increased development costs**

The proposed changes, particularly Option 5, introduce more stringent road requirements aligned with VDOT standards. These infrastructure improvements could increase development costs, potentially leading to higher housing prices or discouraging some developments altogether.

**4. Preservation of agricultural land**

The clustering options and increased lot sizes aim to preserve more agricultural land. While this supports the Comprehensive Plan's goal of maintaining rural character, it may restrict the land available for housing development, potentially increasing land costs for developers.

**5. Potential shift in development location**

By making development in A-1 and RA districts more challenging, these changes may redirect housing development towards areas within the Urban Services Boundary and other identified growth areas, aligning with the Comprehensive Plan's goals. This could lead to more efficient use of existing infrastructure but may also increase competition for developable land in these areas.
:::

This section summarizes the important development regulations for residential subdivision in Culpeper County, and compares them to the set of four potential reforms presented to the Planning Commission by staff in August 2024. These are listed as Options 2 through 5. (Staff used Option 1 to describe a "no change" approach.)

## Division types

County code outlines four different subdivision methods, each with their own rules and guidelines.

According to the Culpeper County Subdivision Ordinance, there are four primary types of divisions. Specifically, Article VI of the Subdivision Ordinance outlines the following types of divisions:

1. **Minor** divisions (Section 611)
2. **10 Acre** divisions (Section 612)
3. **Family** divisions (Section 613)

The ordinance does not explicitly define a "**Major**" division category. However, it does distinguish between Minor divisions and those that require standard procedures (which could be considered "Major" divisions). Furthermore, county staff organized historical subdivision approval data into these four types. (See @sec-dev.)

The tables below outline the main distinctions between the four division types.

| Attribute | Major Division | Minor Division |
|-------------------|----------------|----------------|
| 1. Eligibility Criteria | No specific restrictions; applies when other division types are not applicable | Any existing parcel |
| 2. Number of Lots Created | No specific limit | Up to 3 new lots plus remnant within a 5-year period |
| 3. Minimum Lot Size | As per zoning district requirements | As per zoning district requirements |
| 4. Road Access Requirements | As per VDOT standards | Existing street access |
| 5. Review and Approval Process | Full preliminary and final plat review | Simplified review process |
| 6. Development Standards | Full compliance with all subdivision standards | May be exempted from some standard procedures |
| 7. Time Restrictions | None specified | No more than once every 5 years |
| 8. Special Conditions or Restrictions | None specific to this type | Must not adversely affect remainder of parcel or adjoining property |
| 9. Intended Purpose/Use | Large-scale residential development | Small-scale divisions of land |

<br>

| Attribute | 10 Acre Division | Family Division |
|-------------------|-------------------|-----------------|
| 1. Eligibility Criteria | Only in A-1 (Agricultural) or RA (Rural Residential) districts | In A-1, RA, RR, R-1, or R-2 districts; property owned for at least 5 years |
| 2. Number of Lots Created | No specific limit mentioned | Single division per family member |
| 3. Minimum Lot Size | 10 acres for each lot, including remnant | Minimum of zoning district or 1 acre, whichever is less |
| 4. Road Access Requirements | 50-foot wide easement to public street | 20-foot wide easement to public street |
| 5. Review and Approval Process | Simplified review process | Simplified review process |
| 6. Development Standards | May be exempted from some standard procedures | May be exempted from some standard procedures |
| 7. Time Restrictions | None specified | Grantee cannot transfer for 5 years |
| 8. Special Conditions or Restrictions | Requires perpetual maintenance agreement for private roads | Only for immediate family members; affidavit required |
| 9. Intended Purpose/Use | Large-lot rural development | Allowing property owners to give/sell land to family members |

Major subdivisions follow standard procedures and are suitable for larger-scale developments, while Minor subdivisions offer a more streamlined process for creating up to three new lots every five years. The 10 Acre division option, available only in agricultural and rural residential zones, allows for the creation of large rural lots with fewer regulatory hurdles. Family divisions provide an opportunity for landowners to transfer property to immediate family members with reduced lot size requirements and simplified procedures, but come with restrictions on future transfers.

## Minimum lot size

:pushpin: **Current requirements**

The primary criteria to determine whether a parcel can be subdivided is its lot size. Current minimum lot sizes for each zoning district are shown in the table below.

| District | Minimum Lot Size          | Notes                                        |
|----------|---------------------------|----------------------------------------------|
| A-1      | 5 acres                   | 10 acres for lots on private roads/easements |
| RA       | 3 acres                   | 10 acres for lots on private roads/easements |
| RR       | 3 acres                   |                                              |
| R-1      | 40,000 sqft (0.918 acres) |                                              |
| R-2      | 25,000 sqft (0.574 acres) |                                              |
| R-3      | 15,000 sqft (0.344 acres) | 8,700 sq ft for townhouses                   |
| R-4      | 5 acres                   | For multifamily dwellings                    |
| PUD      | 200 acres                 | Maximum 3 dwelling units per acre            |

<br>
:arrows_counterclockwise: **Proposed requirements**

The proposed options significantly increase minimum lot sizes in the A-1 and RA districts, potentially reducing overall density in these areas. Options 3, 4, and 5 introduce clustering provisions, allowing for smaller lot sizes (1-2 acres) in exchange for preserving larger tracts of land.

| Option | Minimum Lot Size |
|--------|------------------|
| Option 2 | RA: 6 acres<br>A1: 10 or 20 acres |
| Option 3 | A-1: 20 acres<br>Clustered lots: As low as 1 or 2 acres |
| Option 4 | A-1: 20 acres<br>Clustered lots: As low as 1 or 2 acres |
| Option 5 | Clustered lots: 2 acres minimum |

## Density

:pushpin: **Current requirements**

Residential density for new subdivisions in Culpeper County depends on the parcel's zoning.

| Zoning District | Allowable Density                                      |
|-----------------|--------------------------------------------------------|
| A-1             | 1 dwelling unit per 5 acres                            |
| RA              | 1 dwelling unit per 3 acres                            |
| RR              | 1 dwelling unit per 3 acres                            |
| R-1             | 1.09 dwelling units per acre (40,000 sq ft lots)       |
| R-2             | 1.74 dwelling units per acre (25,000 sq ft lots)       |
| R-3 (Single-family) | 2.90 dwelling units per acre (15,000 sq ft lots)       |
| R-3 (Townhouse) | 5 dwelling units per acre                              |
| R-3 (Multifamily) | 8 dwelling units per acre maximum                   |
| R-4             | Varies (multifamily dwellings or condominiums)        |
| PUD             | 3 dwelling units per acre maximum (gross density)      |

<br>
:arrows_counterclockwise: **Proposed requirements**

The proposed options would generally reduce allowable subdivision densities across zoning districts, except when clustering is used.

| Option | Density Allowances |
|--------|---------------------|
| Option 2 | Reduced (due to increased lot sizes) |
| Option 3 | Decreased, but clustering allowed |
| Option 4 | Decreased, but can maintain existing density if clustered |
| Option 5 | Based on current density minus one right for the 70% open space tract |

## Clustering

:pushpin: **Current requirements**

The cluster subdivision requirements in Culpeper County include core standards that apply across all eligible zoning districts, as well as specific requirements that vary by district.

Universal requirements:

1. Minimum development area: 5 acres
2. Conservation area or open space: At least 40% of gross acreage
3. Maximum number of units: No more than total density normally allowable in district

Requirements by zoning district:

| Requirement | RR | R-1 | R-2 | R-3 |
|-------------|----|----|----|----|
| Single-family dwelling lot coverage | ≤ 40% | ≤ 40% | ≤ 40% | ≤ 40% |
| Townhouse unit lot coverage | - | - | - | ≤ 60% |
| Townhouse groupings | - | - | - | ≤ 8 units |
| Minimum lot size (sq ft) | 43,560 | 20,000 | 15,000 | 4,500 |
| Minimum lot width (interior/corner) | 120'/120' | 85'/100' | 75'/90' | 50'/60' |
| Front yard setback | 50' | 45' | 40' | 20' |
| Side yard setback (interior/corner) | 20'/40' | 15'/25' | 10'/20' | 5'/15' |
| Rear yard setback | 35' | 30' | 25' | 15' |

::: {.callout-note}
This table is based on the information provided in the zoning ordinance summary. The A-1 (Agricultural) and RA (Rural Area) districts are not included as they do not have specific cluster subdivision provisions mentioned in the provided documents.
:::

<br>
:arrows_counterclockwise: **Proposed requirements**

The proposed changes introduce clustering options for agricultural and rural areas (A-1 and RA). These new provisions range from allowing clustering to preserve large lots (Option 3), incentivizing it to maintain density (Option 4), to making it a requirement with significant land preservation (Option 5). These outline potential shifts towards more flexible development patterns in rural areas compared to the current fixed-lot approach.

| Option | Clustering Provisions |
|--------|------------------------|
| Option 2 | Not mentioned |
| Option 3 | Allowed, with most permitted dwelling units on smaller lots to preserve a large lot |
| Option 4 | Allowed and incentivized to gain back density |
| Option 5 | Required, with minimum 70% of land deed-restricted to one parcel |

## Road requirements

:pushpin: **Current requirements**

The following table summarizes the current road frontage requirements for residential subdivisions in Culpeper County across different zoning districts. These requirements apply to lots fronting on public streets.

| Zoning District | Minimum Road Frontage (feet) |
|-----------------|------------------------------|
| A-1             | 250                          |
| RA              | 200                          |
| RR              | 200                          |
| R-1             | 120                          |
| R-2             | 100                          |
| R-3             | 80                           |
| R-4             | 200                          |

<br>
:arrows_counterclockwise: **Proposed requirements**

The proposed changes introduce more varied road requirements compared to the current fixed frontage standards for each zoning district. 

| Option | Road Requirements |
|--------|---------------------|
| Option 2 | 250ft frontage for RA to achieve 6-acre lot |
| Option 3 | Could steer clusters on or off state-maintained roads |
| Option 4 | Could steer clusters on or off state-maintained roads |
| Option 5 | Align with VDOT entrance types (PE, LVC, commercial). Standards must be met before building permits issued. |






# Development Analysis {#sec-dev}

From 1993 through 2023, Culpeper County approved approximately 4,569 new lots via subdivision. Most of these were created before the Great Recession, peaking at 728 in 2005. Following a historical low of 37 new lots in 2013, subdivision activity has steadily increased, and is now comparable to the "pre-boom" pace set between 1995 and 2005.

```{r}
#| label: fig-subv-hist
#| fig-cap: "Approved subdivision history (1993-2023)"

subdiv_hist <- read_rds("data/subdiv_hist.rds")

subdiv_hist |> 
  filter(var == "Total") |> 
  ggplot(aes(x = year)) +
  geom_line(aes(y = Lots), color = pal[1]) +
  geom_col(aes(y = Plats), fill = pal[2]) +
  scale_x_continuous(breaks = seq(1993, 2023, 2)) +
  scale_y_continuous(breaks = seq(0, 750, 100), expand = c(0.01,0.05)) +
  add_zero_line() +
  labs(
    title = "Approved subdivision history (1993-2023)",
    subtitle = "Annual number of approved <span style='color:#8baeaa;'>**plats**</span> and <span style='color:#445ca9;'>**lots**</span>",
    caption = "**Source:** *Culpeper County Subdivision History 1993-2023* staff report"
  ) +
  theme_hda() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5)
  )

```

While most of the lots created over this period were the result of Major divisions, most of the actual plats were either Minor or Family divisions.

```{r}
#| label: fig-subv-hist-type
#| fig-cap: "Approved subdivision history by division type"

subdiv_hist |> 
  filter(var != "Total") |> 
  mutate(var = factor(var, levels = c("Major", "Minor", "10 Acre", "Family"))) |> 
  ggplot(aes(x = year)) +
  facet_wrap(~var, axes = "all_x") +
  #geom_text(data = ann_text, aes(x = year, y = Lots), label = "text") +
  geom_line(aes(y = Lots), color = pal[1]) +
  geom_col(aes(y = Plats), fill = pal[2]) +
  coord_cartesian(ylim = c(0, 130)) +
  scale_x_continuous(breaks = seq(1991, 2023, 4)) +
  scale_y_continuous(
    breaks = seq(0, 200, 25),
    expand = c(0.01,0.01)
    ) +
  add_zero_line() +
  labs(
    title = "Approved subdivision history by division type",
    subtitle = "Annual number of approved <span style='color:#8baeaa;'>**plats**</span> and <span style='color:#445ca9;'>**lots**</span>",
    caption = "**Source:** *Culpeper County Subdivision History 1993-2023* staff report<br>**Note:** The 553 Major Division lots approved in 2005 exceed the scale used for this figure."
  ) +
  theme_hda() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5)
  )

```

Building permit approvals for new single-family homes from 2000 through 2023 mirror the trend in subdivision lot approvals. With the exception of the 2008–2010 period, the number of building permits issued per year always exceeded the respective number of new lots created. 

There are two likely reasons for this. First, not all building permits issued were for new construction on subdivision lots. Second, the approval of a new residential lot may come well before the property owner is willing or able to build a home.

```{r}
#| label: fig-subv-bp
#| fig-cap: "Approval history of subdivision lots and building permits"

cbps <- read_rds("data/cbps.rds") |> 
  filter(type == "1-unit")

subdiv_hist_tot <- subdiv_hist |> 
  filter(year >= 2000, var == "Total")

ggplot() +
  geom_line(
    data = subdiv_hist_tot,
    aes(x = year, y = Lots),
    color = pal[1]
    ) +
  geom_line(
    data = cbps,
    aes(x = year, y = bldgs),
    color = pal[5]
    ) +
  scale_x_continuous(breaks = seq(2000, 2023, 2)) +
  scale_y_continuous(
    limits = c(0, 1600),
    breaks = seq(0, 1700, 250),
    expand = c(0.01,0.01)
    ) +
  add_zero_line() +
  labs(
    title = "Approval history of subdivision lots and building permits",
    subtitle = "Annual number of approved <span style='color:#445ca9;'>**lots**</span> and single-family <span style='color:#a97a92;'>**building permits**</span>",
    caption = "**Sources:** *Culpeper County Subdivision History 1993-2023* staff report and Census Building Permits Survey"
  ) +
  theme_hda() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5)
  )

```

# Parcel Analysis

| Description | Count |
| --- | --- |
| Total parcels in GIS data | 18,271 |
| └── Parcels zoned RA or A-1 | 11,803 |
| └─── Not already subdivided lots | 8,435 |
| └──── Not a conservation easement | 8,204 |
| └───── Over 6.0 acres | 4,309 |
| └────── Not in floodplain | 4,290 |

For properties covered only in part by the floodplain, that area is subtracted from the parcel geometry.

Map showing all parcels meeting above criteria:

```{r}
#| label: parcel-map

parcel_filter <- read_rds("gis/parcel_filter.rds")

mapview(
  parcel_filter,
  zcol = "zoning",
  layer.name = "Zoning",
  label = FALSE,
  popup = FALSE
)

```

## Current potential

Number of potential *new* subdivided lots based on current regulations.

```{r}

current <- parcel_filter |> 
  st_drop_geometry() |> 
  filter(
    (zoning == "A1" & acres >= 10) | (zoning == "RA" & acres >= 6)
  ) |> 
  mutate(
    potential_lots = 
      case_when(
        zoning == "A1" ~ floor(acres/5),
        zoning == "RA" ~ floor(acres/3)
    )
  )

current_sum <- current |> 
  summarise(
    parcels = n(),
    lots = sum(potential_lots),
    .by = "zoning"
  ) |> 
  adorn_totals()

current_sum |> 
  mutate(across(2:3, ~ formattable::comma(.x, digits = 0))) |> 
  kable(
    col.names = c("Zoning", "Eligible parcels", "Potential new lots"),
    align = "lrr"
    ) |> 
  kableExtra::kable_styling(bootstrap_options = c("hover", "condensed", "responsive", "striped"))

```

::: {.callout-warning}
### Note
Methods do not yet incorporate road frontage requirements or clustering options.
:::

Methods:

- Filter parcels to at least twice the lot size requirement for its zoning
- Divide acreage by minimum lot size to calculate number of potential lots
- Round down results to whole numbers

```{r}
#| label: fig-current
#| fig-cap: "Most eligible parcels can create 1 or 2 new lots"

current |> 
  mutate(
    bin = fct_case_when(
      potential_lots == 1 ~ "1",
      potential_lots == 2 ~ "2",
      potential_lots == 3 ~ "3",
      potential_lots == 4 ~ "4",
      potential_lots == 5 ~ "5",
      potential_lots >= 6 & potential_lots <= 9 ~ "6-9",
      potential_lots >= 10 & potential_lots <= 19 ~ "10-19",
      potential_lots >= 20 ~ "20+"
    )
  ) |> 
  ggplot(aes(x = bin, fill = zoning)) +
  facet_wrap(~zoning) +
  geom_bar() +
  scale_fill_hda() +
  add_zero_line() +
  labs(
    title = "Most eligible parcels can create 1 or 2 new lots",
    subtitle = "Under current regulations",
    x = "New lots per parcel",
    y = "Parcels"
  ) +
  theme_hda() +
  theme(axis.title = element_text())

```


## Proposed potential

### Option 2 {.unnumbered}

*From staff presentation:*

> **Increase Minimum Lot Sizes (Reduced Density)**
>
> * Raise the lot minimums for both districts, such as 6 acres for the RA and 10 or 20 acres for the A1.
> * Require Road frontage requirements of 250ft. for RA to achieve 6-acre lot.
> * Increases minimum lot size to where agricultural uses could be more easily enjoyed and benefit from local land use tax incentives. Such as 6+ acre minimum for some agricultural uses and 20+ acres for forestry uses.

For this scenario, assume new minimum lot sizes:

- RA: 6 acres
- A1: 10 acres

Total number of potential lots is less than half of the estimate under current regulations.

```{r}

opt2 <- parcel_filter |> 
  st_drop_geometry() |>
  filter(
    (zoning == "A1" & acres >= 20) | (zoning == "RA" & acres >= 12)
  ) |> 
  mutate(
    potential_lots = 
      case_when(
        zoning == "A1" ~ floor(acres/10),
        zoning == "RA" ~ floor(acres/6)
      ),
    potential_lots = pmax(potential_lots - 1, 0)
  )

opt2_sum <- opt2 |> 
  summarise(
    parcels = n(),
    lots = sum(potential_lots),
    .by = "zoning"
  ) |> 
  adorn_totals()

opt2_sum |> 
  mutate(across(2:3, ~ formattable::comma(.x, digits = 0))) |> 
  kable(
    col.names = c("Zoning", "Eligible parcels", "Potential new lots"),
    align = "lrr"
    ) |> 
  kableExtra::kable_styling(bootstrap_options = c("hover", "condensed", "responsive", "striped"))

```

::: {.callout-warning}
### Note
Methods do not yet incorporate road frontage requirements or clustering options.
:::

### Option 3 {.unnumbered}

*From staff presentation:*

> **Decrease allowed Density, Paired With Clustering**
>
> * For example, raise the A1 minimum to 20 acres but allow most of the permitted dwelling units to be on smaller lots to preserve a large lot.
> * The minimum lot size for clustered lots could be as low as 1 or 2 acres.
> * Could have a maximum lot size for clustered lots as well.
> * Could steer clusters on or off state-maintained roads.

Decreased density with clustering:

- Increase A1 minimum lot size to 20 acres
- Large lot preserved, but new lots can be 2 acres each

*Based on staff proposals, assume no change to RA parcels in this scenario.*

```{r}

opt3 <- parcel_filter |> 
  st_drop_geometry() |>
  filter(
    zoning == "A1" & acres >= 22
  ) |> 
  mutate(
    potential_lots = pmax(floor((acres - 20) / 2), 0)
  )

opt3_sum <- opt3 |> 
  summarise(
    parcels = n(),
    lots = sum(potential_lots),
    .by = "zoning"
  )

opt3_sum |> 
  mutate(across(2:3, ~ formattable::comma(.x, digits = 0))) |> 
  kable(
    col.names = c("Zoning", "Eligible parcels", "Potential new lots"),
    align = "lrr"
    ) |> 
  kableExtra::kable_styling(bootstrap_options = c("hover", "condensed", "responsive", "striped"))

```

::: {.callout-warning}
### Note
Estimate for Option 3 does not incorporate road frontage and other potential limiting factors. Due to increased clustering density, number of potential new lots is far greater than current regulations, despite the large increase in minimum lot size.
:::

### Option 4 {.unnumbered}

*From staff presentation:*

> **Decrease allowed Density with Incentivized Clustering option (to gain density back)**
>
> Decrease the development rights allowed, allow for clustering, and allow for additional density if clustered.
>
> * For example, raise the A1 minimum to 20 acres, but allow for the old density of 5 acres per dwelling if all but one of the proposed lots are clustered.
> * This preserves a large lot while still maintaining existing density.
> * The minimum lot size for clustered lots could be as low as 1 or 2 acres.
> * Could have a maximum lot size for clustered lots as well.
> * Could steer clusters on or off state-maintained roads.

```{r}

```


### Option 5 {.unnumbered}

*From staff presentation:*

> **Require clustering in A1/RA but allow option that uses current time-based regulations**
>
> Create a new primary minor subdivision ordinance that requires clustering, with an alternative option that uses the current time-based system and lot minimums.
>
> * Minimum 70% of land must be deed-restricted to one parcel with current rights of a parcel in A1/RA excepting subdivision rights
> * Minimum 70% of any state road frontage must convey with the 70% land if practicable (if it can be achieved without reducing the permitted density in the 30% cluster
>
> 30% of land developable with the following conditions:
> 
> * The number of building rights would be the total density currently allowed less one right for the 70% open space tract.
> * Minimum road standards would be required.
> * Minimum lot size is 2 acres.
> * Will need to possibly consider a soils overlay district. As some parcels in the “shrink-swell” soil areas could realize an increased density with this methodology.

::: {.content-hidden}
```
Application of policy analysis to a detailed spatial analysis of parcel and land use data. Calculation of eligible properties and number of potential lots under current versus proposed regulations. Incorporation of land conservation data and other relevant development limitations to increase accuracy of estimates.

For the Parcel Analysis, HDA will apply the specific attributes and geometric criteria enumerated in the current and proposed subdivision regulations to county parcel records. This process will determine:

* The number of eligible parcels and potential resultant lots under current subdivision regulations
* The number of eligible parcels and potential resultant lots under proposed subdivision regulations

To create more realistic estimates, HDA will incorporate discrete variables that limit the location and intensity of potential residential development on properties that would otherwise be eligible for subdivision. These variables include, but are not limited to:

* Parcels in the Land Use Program or containing conservation easements
* Presence of streams, wetlands, or other natural drainageways
* Flood zones
* Existing or potential road frontage
```
:::

# Findings

```
* The overall scope and significance of the proposed subdivision ordinance changes
* The degree to which subdivisions are limited by factors outside of any current or proposed regulations in the subdivision ordinance, including state requirements for roads and environmental protections
* Whether any increased limitations on residential subdivisions might restrict development activity significantly enough to cap development at or below historical averages
```





